Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-02 demystify.tex
reognizing

@@

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}
% \section{Why the US and not other countries?}
% TODO

\section{How is Your Application Evaluated?}\label{sec:evalapps}

@@

\subsection*{Admission Committee (adcom)}\label{sec:adcom}

@@

\section{Who is this document for?}
\sectioninfo{This document is mainly intended for international students from smaller countries and domestic students from less well-known universities.}

@@

\subsubsection{Why CS depts do not waive the application fee?}

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, most CS programs already receive many applications and would be overwhelmed if the application process were free.  Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.
\end{commentbox}
\subsection{How are decisions made?}\label{sec:how-decisions}
\subsectioninfo{Even if all adcom reviewers recommend acceptance, the application can still be rejected. Vice versa, if all reviewers think the application is weak, the student might still be admitted.}

@@

\subsection{Do adcom members talk to each other?}\label{sec:adcom-discuss}
\subsectioninfo{Sometimes adcom members discuss applicants, but in most cases they make independent decisions.}

@@

\subsection{Do adcom members read all applications?}\label{sec:read-all}
\subsectioninfo{Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications}

@@

\subsection{Timeline}

The timeline for interviews and notification letters varies.  Faculty set up interviews based on their (busy and erratic) schedule. Do not be surprised if you get an interview invitation at the last minute. Some schools do not do interviews at all (\autoref{sec:no-interview}).

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their waitlist).  In general, you should hear back from most schools by mid-March (but some schools might take longer!). Note that rejection letters might come very late (see \autoref{sec:late-rejection}).

In short, there is not much you can do about this other than to be patient and wait. Also do not send emails asking about interviews or status (unless you have interviewed specifically with someone then you can ask that person for status updates and others \autoref{sec:accept-postpone-decline}).

@@

@@

\chapter{Getting Admitted}\label{sec:accepted}

\chapterinfo{Congrats! Now it is your turn to evaluate the school!
Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now.}

@@

If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for stronger LoRs, etc. See rejection reasons (\autoref{sec:why-rejected}) and timeline (\autoref{sec:late-rejection}), and additional advice to improve your chance next time (\autoref{sec:improve-your-chance}).

@@

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  Profs. might not care much about these, but the college or the university often sets a minimum that you need to pass to be considered, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while profs. are willing to argue for your case, they might be reluctant to go against the requirement of the university on standard exams.

@@

\subsection{Timeline: Why are rejection letters sent so late?}\label{sec:late-rejection}
\subsectioninfo{Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

\paragraph{Rejection Letters} Schools typically start sending out rejection letters to remaining applicants \emph{after they have finalized their admissions decisions}. Thus, rejection letters are often sent out late (e.g., after April 15th or even much later). Not much you can do here, you can try to contact the school to ask about your status, but they might not reply, they might say they are still reviewing applications, or give you inaccurate information (e.g., you will hear in two weeks). In short, you just have to be patient and wait (however, beware that some schools do not send out rejection letters at all).

\subsubsection{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We are not willing (sometimes not allowed) to reveal your evaluation results or give you feedback on how to improve your profile. \emph{So just move on}.  If you really want advice, ask your professors, collaborators, ref writers, or those who have previously applied.

@@

\chapter{Funding}\label{sec:funding}

\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs are provided by the department to help with classes. RAs are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but can be very competitive.}

>>>>

\section{Who is this document for?}
\sectioninfo{This document is intended for international students from smaller countries and domestic students from less well-known universities.}

@@

% \begin{table}[h!]
% \centering
% \begin{tabular}{l|l}
%     \toprule
% \textbf{Month} & \textbf{Task} \\
% \midrule
% January-March & Research programs and prepare for tests \\
% April-June & Take standardized tests (GRE, TOEFL, etc.) \\
% July-September & Prepare application materials (SOP, LORs, etc.) \\
% October-December & Submit applications \\
% January-March & Interviews and follow-ups \\
% April & Admission decisions \\
% \bottomrule
% \end{tabular}
% \caption{Application Timeline}
% \end{table}
% \section{Why the US and not other countries?}
% TODO
\subsubsection{Why CS depts do not waive the application fee?}

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, most CS programs already receive many applications and would be overwhelmed if the application process were free.  Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.
\end{commentbox}

\chapter{How is Your Application Evaluated?}\label{sec:evalapps}

@@

\section{Admission Committee (adcom)}\label{sec:adcom}

@@

@@

\section{How are decisions made?}\label{sec:how-decisions}
\sectioninfo{Even if all adcom reviewers recommend acceptance, the application can still be rejected. Vice versa, if all reviewers think the application is weak, the student might still be admitted.}

@@

\section{Do adcom members talk to each other?}\label{sec:adcom-discuss}
\sectioninfo{Sometimes adcom members discuss applicants, but in most cases they make independent decisions.}

@@

\section{Do adcom members read all applications?}\label{sec:read-all}
\sectioninfo{Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications}

@@

\paragraph{When do interviews happen?}

The timeline for interviews varies.  Faculty set up interviews based on their (busy and erratic) schedule. Do not be surprised if you get an interview invitation at the last minute. Some schools do not do interviews at all (\autoref{sec:no-interview}).

@@

\subsection{Notification Timeline: Why rejection letters are sent so late?}\label{sec:late-rejection}
\subsectioninfo{Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

Some schools send out admission letters in batches, some do rolling admission, and some do not send anything out (e.g., you're on their waitlist). You should hear back from most schools by mid-March (though rejection can come out very late).

Not much you can do other than to be patient and wait. Do not send emails asking about interviews or status (unless you have interviewed specifically with someone (\autoref{sec:interview}) then you can ask that person for status updates and other questions (\autoref{sec:accept-postpone-decline}).

\paragraph{Rejection Letters} Schools typically start sending out rejection letters to remaining applicants \emph{after they have finalized their admissions decisions}. Thus, rejection letters are often sent out late (e.g., after April 15th or even much later). Not much you can do here. You can try to contact the school to ask about your status, but they might not reply, they might say they are still reviewing applications, or give you inaccurate information (e.g., you will hear in two weeks). In short, you just have to be patient and wait (however, beware that some schools do not send out rejection letters at all).

@@

\chapter{Getting Admitted}\label{sec:accepted}
\chapterinfo{Congrats! Now it is your turn to evaluate the school!
Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now.}

@@

If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for stronger LoRs, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice to improve your chance next time (\autoref{sec:improve-your-chance}).

@@

\paragraph{Should I ask for feedback?}
No, don't bother.  You will likely not get any useful feedback.  We are not willing (sometimes not allowed) to reveal your evaluation results or give you feedback on how to improve your profile. \emph{So just move on}.  If you really want advice, ask your professors, collaborators, ref writers, or those who have previously applied.

@@

\paragraph{Low English exam scores (e.g., IELTS or TOEFL)}  Profs. might not care much about these, but the college or the university often sets a minimum that you need to pass to be considered, especially for TA funding (e.g., you need to communicate well with students as a TA).  Thus, while profs. are willing to argue for your case, they might be reluctant to go against the requirement of the university.

@@

@@

\chapter{Funding}\label{sec:funding}
\chapterinfo{TAs, RAs, and fellowships are main funding sources for PhDs.  TAs are provided by the department to help with classes. RAs are given by profs. to help with their research.  Fellowships, provided by the university, department, or external sources such as government or industry, give move flexibility but can be very competitive.}
